Concurrent Programming with Harmony

### 2020-07-19 book/paper.tex
Latest
Updated some appendices
Added an exercise to the interrupts chapter

It will often be necessary to write code that is both interrupt-safe \emph{and}
thread-safe.  As you might expect, this involves both managing locks and
interrupt levels.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{$e$ == $e$}, \texttt{$e$ != $e$} & two Harmony values of the same type \\
\hline
\texttt{$e$ < $e$, $e$ <= $e$} & any two Harmony values \\
\hline
\texttt{$e$ > $e$, $e$ >= $e$} & any two Harmony values \\
\hline
\texttt{$e$ and $e$ and ...} & two or more booleans \\
\hline
\texttt{$e$ or $e$ or ...} & two or more booleans \\
\hline
\texttt{not $e$} & a boolean \\
\hline
\texttt{-$e$} & an integer \\
\hline
\texttt{$e$ + $e$ + ...} & two or more integers, sets, or lists \\
\hline
\texttt{$e$ - $e$} & two integers or sets \\
\hline
\texttt{$e$ * $e$ * ...} & two or more integers or sets \\
\hline
\texttt{$e$ / $e$} & two integers (integer division) \\
\hline
\texttt{$e$ \% $e$} & two integers (division remainder) \\
\hline
\texttt{$e$~..~$e$} & two integers \\
\hline
\texttt{$e$ [not] in $e$} & first is any Harmony value, second is set \\
\hline
\texttt{$e$ if $e$ else $e$} & middle $e$ must be a boolean \\
\hline
\texttt{\^{}$e$} & an address \\
\hline
\texttt{hash $e$} & any Harmony value \\
\hline
\texttt{choose $e$} & a set \\
\hline
\texttt{min $e$, max $e$} & a set \\
\hline
\texttt{cardinality $e$} & a set \\
\hline
\texttt{keys $e$, len $e$} & a dictionary (includes tuple or list) \\
\hline
\texttt{bagsize $e$} & a bag \\
\hline
\texttt{atLabel $e$} & atom (corresponding to a label) \\
\hline
\texttt{nametag $e$} & $e$ must be \texttt{()} \\
\hline
\texttt{processes $e$} & $e$ must be \texttt{()} \\
\hline
\texttt{\&}\textit{lv} & \textit{lv} is an lvalue \\
\hline
\texttt{stop} \textit{lv} & \textit{lv} is an lvalue \\
\hline
\end{tabular}
\vspace{1em}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
Address $n$ & compute address from $n$ components \\
\hline
Apply & pop $m$ and $i$ and apply $i$ to $m$, pushing a value \\
\hline
Assert & pop $b$ and check that it is \texttt{True} \\
\hline
AtomicInc & increment the atomic counter of this context \\
\hline
AtomicDec & decrement the atomic counter of this context \\
\hline
Continue & no-op (but causes a context switch) \\
\hline
Choose & choose an element from the set on top of the stack \\
\hline
Del [$v$] & delete shared variable $v$ \\
\hline
DelVar [$v$] & delete process variable $v$ \\
\hline
Dict & pop $n$ and $n$ key/value pairs and push a dictionary \\
\hline
Dup & duplicate the top element of the stack \\
\hline
Frame \texttt{m}($a$) & start method \texttt{m} with arguments $a$,
initializing variables.  \\
\hline
Go & pop context and value, push value on context's stack, and add to context bag \\
\hline
Jump $p$ & set program counter to $p$ \\
\hline
JumpCond $e$ $p$ & pop expression and, if equal to $e$, set program counter to $p$ \\
\hline
Load [$v$] & push the value of a shared variable onto the stack \\
\hline
LoadVar [$v$] & push the value of a process variable onto the stack \\
\hline
Move $i$ & move stack element at offset $i$ to top of the stack \\
\hline
$n$-ary \textit{op} & apply $n$-ary operator \textit{op} to the top $n$ elements on the stack \\
\hline
Pop & pop a value of the stack and discard it \\
\hline
Push $c$ & push constant $c$ onto the stack \\
\hline
PushAddress $v$ & push the address of variable $v$ \\
\hline
Return & pop return address, push \texttt{result}, and restore program counter \\
\hline
Set & pop $n$ and $n$ elements and push a set of the elements \\
\hline
Spawn & pop tag, argument, and method and spawn a new context \\
\hline
Split & pop set and push minimum and remainder \\
\hline
Stop [$v$] & save context into shared variable $v$ and remove from context bag \\
\hline
Store [$v$] & pop a value from the stack and store it in a shared variable \\
\hline
StoreVar [$v$] & pop a value from the stack and store it in a process variable \\
\hline
\end{tabular}
\vspace{1em}

Clarifications:
\begin{itemize}
\item The \texttt{Address} instruction expects $n$ values $v_1, ..., v_n$ on the stack.
The top value must be an address value.  The remaining values are keys into dictionaries.
The instruction then computes the address of \texttt{\^}$v_1~v_2~...~v_n$.
\item Even though Harmony code does not allow taking addresses of process variables, both
shared and process variables can have addresses.
\item The \texttt{Load}, \texttt{LoadVar}, \texttt{Del}, \texttt{DelVar},
and \texttt{Stop} instructions have an optional
variable name: if omitted the top of the stack must contain the address of
the variable.
\item \texttt{Store} and \texttt{StoreVar} instructions have an optional
variable name.  In both cases the value to be assigned is on the top
of the stack.  If the name is omitted, the address is underneath that
value on the stack.
\item The effect of the \texttt{Apply} instructions depends much on $m$.
If $m$ is a dictionary, then \texttt{Apply} finds $i$ in the dictionary
and pushes the value.
If $m$ is a program counter, then \texttt{Apply} invokes method $m$ by
pushing the current program counter and setting the program counter to
$m$.  $m$ is supposed to leave the result on the stack.
\item The \texttt{Frame} instruction pushes the value of the process
register (\emph{i.e.}, the values of the process variables) onto the
stack.  It initializes the \texttt{result} variable to the empty dictionary.
The \texttt{Return} instruction restores the process register by popping
its value of the stack.
\item All method calls have exactly one argument, although it sometimes
appears otherwise:
\begin{itemize}
\item \texttt{m()} invokes method \texttt{m} with the empty dictionary \texttt{()} as argument;
\item \texttt{m($a$)} invokes method \texttt{m} with argument $a$;
\item \texttt{m($a$, $b$, $c$)} invokes method \texttt{m} with tuple \texttt{($a$, $b$, $c$)} as argument.
\end{itemize}
The \texttt{Frame} instruction unpacks the argument to the method and places them into process variables by the given names.
\item Every \texttt{Stop} instruction must immediately be followed by a 
\texttt{Continue} instruction.
\end{itemize}

>>>>

\begin{figure}
\begin{code}
\harmonysource{trap6}
\end{code}
\caption{\harmonylink{code/trap6.hny} Code that is both interrupt-safe and thread-safe.}
\label{fig:trap6}
\end{figure}

It will often be necessary to write code that is both interrupt-safe \emph{and}
thread-safe.  As you might expect, this involves both managing locks and
interrupt levels.
To increment \textit{count}, the interrupt level must be \textit{True} and
\textit{countlock} must be held.
\autoref{fig:trap6} gives an example of how this might be done.
The important rule to remember is for a process to disable interrupts \emph{before}
attempting to acquire a lock.

Try moving \texttt{lock}() to the beginning of \texttt{increment}()
and \texttt{unlock}() to the end of \texttt{increment}() and see what happens.
While Harmony will only report one faulty run, this incorrect code can lead to
the assertion failing as well as processes getting blocked indefinitely.

\section*{Exercises}
\begin{problems}
\item Make methods \texttt{produce}() and \texttt{consume}() in \autoref{fig:boundedbuffer}
both thread-safe and interrupt-safe.
\end{problems}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{$e$ == $e$}, \texttt{$e$ != $e$} & two Harmony values of the same type \\
\hline
\texttt{$e$ < $e$, $e$ <= $e$} & any two Harmony values \\
\hline
\texttt{$e$ > $e$, $e$ >= $e$} & any two Harmony values \\
\hline
\texttt{$e$ and $e$ and ...} & two or more booleans \\
\hline
\texttt{$e$ or $e$ or ...} & two or more booleans \\
\hline
\texttt{not $e$} & a boolean \\
\hline
\texttt{-$e$} & an integer \\
\hline
\texttt{$e$ + $e$ + ...} & two or more integers, sets, or lists \\
\hline
\texttt{$e$ - $e$} & two integers or sets \\
\hline
\texttt{$e$ * $e$ * ...} & two or more integers or sets \\
\hline
\texttt{$e$ / $e$} & two integers (integer division) \\
\hline
\texttt{$e$ \% $e$} & two integers (division remainder) \\
\hline
\texttt{$e$~..~$e$} & two integers \\
\hline
\texttt{$e$ [not] in $e$} & first is any Harmony value, second is set \\
\hline
\texttt{$e$ if $e$ else $e$} & middle $e$ must be a boolean \\
\hline
\texttt{\^{}$e$} & an address \\
\hline
\texttt{choose $e$} & a set \\
\hline
\texttt{min $e$, max $e$} & a set \\
\hline
\texttt{cardinality $e$} & a set \\
\hline
\texttt{keys $e$, len $e$} & a dictionary (includes tuple or list) \\
\hline
\texttt{bagsize $e$} & a bag \\
\hline
\texttt{atLabel $e$} & atom (corresponding to a label) \\
\hline
\texttt{nametag $e$} & $e$ must be \texttt{()} \\
\hline
\texttt{processes $e$} & $e$ must be \texttt{()} \\
\hline
\texttt{setintlevel} \textit{e} & \textit{e} is a boolean \\
\hline
\texttt{stop} \textit{lv} & \textit{lv} is an lvalue \\
\hline
\texttt{\&}\textit{lv} & \textit{lv} is an lvalue \\
\hline
\end{tabular}
\vspace{1em}

@@

{\small
\begin{tabular}{|l|l|}
\hline
Address $n$ & compute address from $n$ components \\
\hline
Apply & pop $m$ and $i$ and apply $i$ to $m$, pushing a value \\
\hline
Assert & pop $b$ and check that it is \texttt{True} \\
\hline
AtomicInc & increment the atomic counter of this context \\
\hline
AtomicDec & decrement the atomic counter of this context \\
\hline
Continue & no-op (but causes a context switch) \\
\hline
Choose & choose an element from the set on top of the stack \\
\hline
Del [$v$] & delete shared variable $v$ \\
\hline
DelVar [$v$] & delete process variable $v$ \\
\hline
Dict & pop $n$ and $n$ key/value pairs and push a dictionary \\
\hline
Dup & duplicate the top element of the stack \\
\hline
Frame \texttt{m}($a$) & start method \texttt{m} with arguments $a$,
initializing variables.  \\
\hline
Go & pop context and value, push value on context's stack, and add to context bag \\
\hline
Jump $p$ & set program counter to $p$ \\
\hline
JumpCond $e$ $p$ & pop expression and, if equal to $e$, set program counter to $p$ \\
\hline
Load [$v$] & push the value of a shared variable onto the stack \\
\hline
LoadVar [$v$] & push the value of a process variable onto the stack \\
\hline
Move $i$ & move stack element at offset $i$ to top of the stack \\
\hline
$n$-ary \textit{op} & apply $n$-ary operator \textit{op} to the top $n$ elements on the stack \\
\hline
Pop & pop a value of the stack and discard it \\
\hline
Push $c$ & push constant $c$ onto the stack \\
\hline
PushAddress $v$ & push the address of variable $v$ \\
\hline
Return & pop return address, push \texttt{result}, and restore program counter \\
\hline
Set & pop $n$ and $n$ elements and push a set of the elements \\
\hline
SetIntLevel & pop $e$, set interrupt level to $e$, and push old interrupt level \\
\hline
Spawn & pop tag, argument, and method and spawn a new context \\
\hline
Split & pop set and push minimum and remainder \\
\hline
Stop [$v$] & save context into shared variable $v$ and remove from context bag \\
\hline
Store [$v$] & pop a value from the stack and store it in a shared variable \\
\hline
StoreVar [$v$] & pop a value from the stack and store it in a process variable \\
\hline
Trap & pop interrupt argument and method \\
\hline
\end{tabular}
}

\newpage
Clarifications:
\begin{itemize}
\item The \texttt{Address} instruction expects $n$ values $v_1, ..., v_n$ on the stack.
The top value must be an address value.  The remaining values are keys into dictionaries.
The instruction then computes the address of \texttt{\^}$v_1~v_2~...~v_n$.
\item Even though Harmony code does not allow taking addresses of process variables, both
shared and process variables can have addresses.
\item The \texttt{Load}, \texttt{LoadVar}, \texttt{Del}, \texttt{DelVar},
and \texttt{Stop} instructions have an optional
variable name: if omitted the top of the stack must contain the address of
the variable.
\item \texttt{Store} and \texttt{StoreVar} instructions have an optional
variable name.  In both cases the value to be assigned is on the top
of the stack.  If the name is omitted, the address is underneath that
value on the stack.
\item The effect of the \texttt{Apply} instructions depends much on $m$.
If $m$ is a dictionary, then \texttt{Apply} finds $i$ in the dictionary
and pushes the value.
If $m$ is a program counter, then \texttt{Apply} invokes method $m$ by
pushing the current program counter and setting the program counter to
$m$.  $m$ is supposed to leave the result on the stack.
\item The \texttt{Frame} instruction pushes the value of the process
register (\emph{i.e.}, the values of the process variables) onto the
stack.  It initializes the \texttt{result} variable to the empty dictionary.
The \texttt{Return} instruction restores the process register by popping
its value of the stack.
\item All method calls have exactly one argument, although it sometimes
appears otherwise:
\begin{itemize}
\item \texttt{m()} invokes method \texttt{m} with the empty dictionary \texttt{()} as argument;
\item \texttt{m($a$)} invokes method \texttt{m} with argument $a$;
\item \texttt{m($a$, $b$, $c$)} invokes method \texttt{m} with tuple \texttt{($a$, $b$, $c$)} as argument.
\end{itemize}
The \texttt{Frame} instruction unpacks the argument to the method and places them into process variables by the given names.
\item Every \texttt{Stop} instruction must immediately be followed by a 
\texttt{Continue} instruction.
\end{itemize}
